---
title: "Final Project"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r libraries}
library(tidyverse)
library(readxl)

```

```{r import data}
group_raw <- read_excel("data for each group_2025.xlsx", 
                        sheet = 'group')

data_raw <- read_excel("data for each group_2025.xlsx",
                       sheet = 'Raw Data Reformatted',
                       skip = 1) |>
            rename("plot" = 1, "landuse" = 2, "year" = 3) |>
  mutate(year = as_factor(year)) |>
  janitor::clean_names()
```

```{r clean_groups}
#### TIDY GROUP DATA ####
group <- group_raw |>
  ## clean multiple values per cell
  separate_longer_delim(cols = Cropland_2025,
                        delim = ' ') |>
  separate_longer_delim(cols = Grassland_2025,
                        delim = ' ') |>
  separate_longer_delim(cols = Forest_2025,
                        delim = ' ') |>
  separate_longer_delim(cols = Grassland_2024,
                        delim = ' ') |>
  separate_longer_delim(cols = Forest_2024,
                        delim = ' ') |>
  pivot_longer(cols = (contains('_')),
               names_to = 'landuse_year',
               values_to = 'plot') |>
  separate(landuse_year, c('landuse', 'year'), '_') |>
  filter(!is.na(plot), plot != 'plot', plot != '') |>
  distinct() |>
  # mutate(year = as_factor(year)) |>
  janitor::clean_names()

rm(group_raw)

group |>
  filter(group_number == 6)

```

#### SOIL PROPERTIES
convert carbon and nitrogen to mg/kg

convert GMC to percent (already in percent?)

convert bulk density to g/cm3

```{r soil_properties}
soil_properties <- 
  data_raw |>
  select(1:3, total_c_percent:bulk_density_g_cm3) |>
  rename('c_pct' = 4, 
         'n_pct' = 5, 
         'n_' = 6, 
         'pH_KCl' = 7, 
         'mass_soilDry' = 10, 
         'gravimetric_moisture_content' = grametric_moisture_content) |>
  select(-bulk_density_g_cm3, -grametric_moisture_content_percent, -n_)

soil_properties <- 
  soil_properties |>
  mutate(
    bulk_density = mass_soilDry / (3.14*(5/2)^2), # g/cm^3
    n = n_pct * 1000 / 100, #mg/kg
    c = c_pct * 1000 / 100 #mg/kg
    )
```

-   ***c_pct:*** \[GIVEN\] Percentage of soil mass that is carbon (grams carbon/grams soil). Value between 0 and 100

-   ***n_pct:*** \[GIVEN\] Percentage of soil mass that is nitrogen (grams nitrogen/grams soil). Value between 0 and 100

-   ***c:*** mass ratio between nitrogen and the total soil sample (mg/kg)

-   ***n:*** mass ratio between carbon and the total soil sample (mg/kg)

-   ***pH_KCl:*** \[GIVEN\] pH-KCl

-   ***gravimetric_moisture_content:*** \[GIVEN\] the mass of water in the soil divided by the mass of the dry material, typically expressed as a percentage or decimal (e.g., g water/g dry soil). It's measured by weighing a wet sample, oven-drying it until all water evaporates, and then re-weighing the dry sample; the difference reveals the water mass.

-   ***mass_soilDry:*** \[GIVEN\] the dry mass of a 5cm deep and 5cm diameter soil core

-   ***mass_soilWet:*** the wet mass of a 5cm deep and 5cm diameter soil core

-   ***bulk_density*** the average density of the soil core

#### WORMS
convert to individuals per m2

```{r worms}

worms <- 
  data_raw |>
  select(1:3, earthworm_number_individuals_0_12_m2:earthworm_density_g_m2)


worms <- 
  worms |>
  mutate(
    worms_per_m2 = earthworm_number_individuals_0_12_m2 / 0.12, 
    wormMass_per_m2 = earthworm_fresh_weight_g_0_12m2 / 0.12 #g/m^2
         ) |>
  select(plot, landuse, year, worms_per_m2, wormMass_per_m2)
```

-   ***bulk_density*** the average density of the soil core


#### DECOMPOSITION
convert data to percent (already in percent?)
```{r}
decomp <- data_raw |>
  select(1:3, material_loss_after_34_days_percent_16:decompostion_constant_k_d_1) |>
  mutate(material_loss = material_loss_after_34_days_percent_16) |>
  select(plot, landuse, year, material_loss, 'decomp_constant' = decompostion_constant_k_d_1)
  
```

#### ARTHROPODS
convert to individuals per square meter
```{r}

arthropods <- data_raw |>
  select(1:3, total_microarthropods_individuals_per_soil_core:other_microarthropods_ind_m2)

names(arthropods) <- gsub("s_individuals_per_soil_core", 
                         "_count", 
                         colnames(arthropods))
arthropods <- 
  arthropods |> 
  rename('microarthopod_count' = total_microarthropod_count,
         'other_count' = other_microarthropod_count) |>
  mutate(
    microarthopod_density = microarthopod_count / (5/2)^2,
    mite_density = mite_count / (5/2)^2,
    springtail_density = springtail_count / (5/2)^2,
    other_density = other_count / (5/2)^2
  ) |>
  select(1:3, contains('count'), contains('density'))
```

#### ENZYMES

convert data to µg PNP/g dry soil per hour

gmc = (wet-dry) / dry
dry = wet / (gmc +1)

we assume there is no interaction between soil moisture and pnp production
```{r}
enzymes <- data_raw |>
  select(1:3, 27:28)

enzymes <-
  enzymes |>
  left_join(soil_properties |>
              select(1:3, 'gmc' = gravimetric_moisture_content)) |>
  rename('pnp_wet' = 4, 
         'pnp_dry' = 5) |>
  mutate(dry = 1 / (gmc + 1), 
         pnp_dry = pnp_wet / dry)
  
```
-   ***pnp_wet:*** \[GIVEN\] amount of pnp produced in micrograms per gram of fresh soil

-   ***dry:*** amount of dry soil in one gram of fresh soil

-   ***pnp_dry:*** amount of pnp produced in micrograms per gram of dry soil

#### NEMATODES

Task: convert data to ind./100 g dry soil

gmc = (wet-dry) / dry
dry = wet / (gmc +1)

```{r nematodes}
nematodes <- data_raw |>
  select(1:3, 29:30)

# nematodes |>
#   rename('nematode_count' = 4) |>
#   mutate(nematode_estimate_100 = nematode_count * 2)

nematodes <-
  nematodes |>
  rename('nematode_count' = 4) |>
  left_join(soil_properties |>
              select(1:3, 'gmc' = gravimetric_moisture_content)) |>
  mutate(
    dry = 50 / (gmc + 1),
    nematode_density_dry = nematode_count / dry,
    nematode_estimate_100 = nematode_density_dry * 100
  ) |>
  select(-nematodes_100_g_dry_soil) |>
  glimpse()
```

-   ***nematode_count:*** \[GIVEN\] number of nematodes observed in 50g of fresh soil. *strangely, it contains partial numbers.*

-   ***gmc:*** \[GIVEN\] gravimetric_moisture_content

-   ***dry:***  amount of dry soil in the fresh soil sample

-   ***nematode_density_dry:*** number of nematodes expected to be found in 1g of dry soil.

-   ***nematode_estimate_100:*** number of nematodes expected to be found in 100g of soil

#### MCIROBIAL BIOMASS CARBON

calculate mbc according to the protocol

special note: fumigated should be higher than non-fumigated. If it is not, swap the values.

```{r}
mbc <- data_raw |>
  select(1:3, 31:32, 34:37)

mbc <-
mbc |>
  left_join(soil_properties |> select(1:3, gravimetric_moisture_content)) |> 
  rename(
    'soil_wet_weight' = mass_fresh_soil_g,
    'c_fumigated' = fumigated_c_mg_l,
    'c_nonfumigated' = non_fumigated_c_mg_l) |>
  mutate(
    extractant_volume = 60, #mL
    
    ws = gravimetric_moisture_content, #g water / g dry soil
    
    ms = soil_wet_weight / (1+ws),
    
    vs = (soil_wet_weight - ms) + extractant_volume,
    
    cf = c_fumigated * vs / ms,
    cnf = c_nonfumigated * vs / ms,
    
    mbc = (cf-cnf) / .045 # 0.45 is k_ec value
  )
```


#### EXPLORATORY DATA ANALYSIS

```{r ph_decomposition}
left_join(decomp, soil_properties) |>
  ggplot() +
  geom_point(aes(x = pH_KCl, y = material_loss, color = landuse))
```

```{r ph_landuse}
soil_properties |>
  ggplot() +
  geom_boxplot(aes(x = landuse, y = pH_KCl, fill = landuse))
```

```{r ph_population}

```

```{r landuse_populations}
left_join(nematodes, arthropods) |> 
  left_join(soil_properties) |>
  pivot_longer(cols = contains('density'),
               names_to = 'organism',
               values_to = 'density') |>
  ggplot() +
  geom_boxplot(aes(x = organism, y = density)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~landuse)
  glimpse()
```

```{r bulkDensity_landuse}
soil_properties |>
  ggplot() +
  geom_boxplot(aes(x = landuse, y = bulk_density, fill = landuse))
```

```{r enzymeActivity_soilCarbon}
left_join(soil_properties, enzymes) |>
    # glimpse()
  ggplot() +
  geom_point(aes(x = c, y = pnp_dry, color = landuse))

```

